---
title: "Transcripts with high distal heritability mediate genetic effects on complex traits"
output: 
  pdf_document:
    keep_tex: true
    includes:
      in_header: header_simple.tex
---

```{r load_data, echo = FALSE, message = FALSE, warning = FALSE}
library(here)
pheno.stats <- readRDS(here("Data", "pheno_stats.RDS"))
comp.trait.herit <- readRDS(here("Data", "Composite_Trait_Heritability.RDS"))
comp.trait.exp <- readRDS(here("Data", "Model_Variance_Explained.RDS"))
local.distal.comp <- readRDS(here("Data", "Local_Distal_Comparison.RDS"))
min.herit <- floor(min(sapply(lapply(local.distal.comp, function(x) apply(x, 2, median, na.rm = TRUE)), min))*100)
max.herit <- ceiling(max(sapply(lapply(local.distal.comp, function(x) apply(x, 2, median, na.rm = TRUE)), max))*100)
var.pcor <- readRDS(here("Data", "Pcor.RDS"))
lit.enrich <- readRDS(here("Data", "Lit_enrich.RDS"))
```

```{r pfun, echo = FALSE}
#we don't want to display p values less than 2.2e-16
#use this function to threshold p values
threshold_p <- function(p, thresh = 2.2e-16){
  if(p < thresh){
    return(thresh)
  }else{
    return(p)
  }
}
```

```{r supp_fig_tracker, echo = FALSE}
#we need to keep track of supplemental figure numbers
#without having captions. We want to create a reference
#number for each figure that relates to the order they
#appear in the manuscript, but remains the same if the
#figure is referenced more than once.

#I pulled this from the epigenetics ms. Update for this
#paper

#create a vector of reference numbers for each figure
sup_labels <- c("fig:eQTL", "fig:top_enrich_kegg", 
  "fig:cc_imputation", "fig:kegg_highlights", "fig:Nucb2_eqtl")
fig.counter <- rep(0, length(sup_labels))
names(fig.counter)  <- sup_labels

ref_fig <- function(fig.vector, fig.ref){
  if(fig.vector[fig.ref] == 0){
    fig.vector[fig.ref] <- max(fig.vector)+1
  }
  return(fig.vector)
}
```

Anna L Tyler, J Matthew Mahoney, Mark P Keller, Candice N. Baker, Anuj Srivastava,
Isabela Gerdes Gyuricza, [Other Churchill people? Jeff Harder?], Nadia A Rosenthal,
Alan D Attie, Gary A Churchill and Gregory W Carter

## Abstract
Gene expression is an important mediator of genetic effects on phenotype. 
Although many genes are subject to simple, local regulation, recent evidence 
suggests that complex distal regulation may be more important to trait variability. 
To investigate this possibility, we combined two large, purpose-built data sets 
modeling diet-induced obesity and metabolic disease in genetically diverse 
mice. Using a novel high-dimensional mediation analysis, we identified a 
heritable composite transcript that explained 30% of the variation across all
metabolic traits. The composite transcript was interpretable in terms of 
enriched biological processes and predicted obesity status in an independent 
mouse cohort as well as human cohorts with measured gene expression. 
Transcripts contributing most strongly to this composite mediator tended 
to have complex, distal regulation distributed throughout the genome. 
These results suggest that trait-relevant variation in transcription is largely 
distally regulated, but is nonetheless identifiable, interpretable, and translatable 
across species.





## Introduction

In the quest to understand the genetic architecture of complex traits, 
gene expression is an important mediator between genotype and phenotype. 
There is ample evidence from genome-wide association studies 
(GWAS) that regulation of gene expression accounts for the bulk 
of the genetic effect on complex traits, as most trait-associated 
variants lie in gene regulatory regions \cite{pmid22955828, 
pmid25363779, pmid21617055, pmid19474294, pmid24702953, 
pmid24316577, pmid27126046}. It is widely assumed that these 
variants influence local transcription, and methods such as 
transcriptome-wide association studies (TWAS) 
\cite{pmid33020666, pmid26258848, pmid27019110, pmid26854917},
summary data-based Mendelian randomization (SMR) \cite{pmid27019110}, 
and others capitalize on this idea to identify genes associated 
with multiple disease traits \cite{pmid29567659, pmid35533209, pmid27309819, 
pmid30950127}

Despite the great promise of these methods, explaining trait 
effects with local gene regulation has been more difficult than
initially assumed \cite{pmid32912663, pmid36515579}. Although 
trait-associated variants tend to lie in non-coding, regulatory 
regions, they often do not have detectable effects on gene expression 
\cite{pmid32912663} and tend not to co-localize with expression 
quantitative trait loci (eQTLs) \cite{pmid36515579, pmid37857933}. 

One possible explanation for these observations is that gene 
expression is not being measured in the appropriate cell types 
and thus local eQTLs influencing traits cannot be detected 
\cite{pmid32912663}. An alternative explanation that has been 
discussed in recent years is that effects of these variants 
are mediated not through local regulation of gene expression, 
but through distal regulation \cite{pmid37857933, pmid32424349, 
pmid32831138, pmid30950127}. 

In this model, a gene's expression is influenced by 
many variants throughout the genome through their cumulative 
effects on a broader regulatory network. In other 
words, the heritable component of the transcriptome is an emergent 
state arising from the myriad molecular interactions defining and 
constraining gene expression. 

To assess the role of wide-spread distal gene regulation 
on complex traits, we investigated diet-induced obesity and 
metabolic disease as an archetypal example. Diet-induced 
obesity and metabolic disease are genetically complex with 
hundreds of variants mapped through GWAS [REFS]. These variants 
are known to act through multiple tissues that interact dynamically
with each other [REFS], including 
adipose tissue, pancreatic islets, liver, and skeletal muscle. 
The multi-system 
etiology of metabolic disease complicates mechanistic dissection 
of the genetic architecture, requiring large, dedicated data sets 
that include high-dimensional, clinically relevant phenotyping, dense 
genotyping in a highly recombined population, and transcriptome-wide 
measurements of gene expression in multiple tissues. Measuring gene 
expression in multiple tissues is critical to adequately assess the extent 
to which local gene regulation varies across multiple tissues and whether 
such variablilty might account for previous failed attempts to identify 
trait-relevant local eQTL. Such data sets are extremely difficult to 
obtain in human populations, particularly in the large numbers of 
subjects required for adequate statistical power. Thus, to investigate 
further the role of local and distal gene regulation on complex traits, 
we have generated an appropriate data set in a large population of 
diversity outbred (DO) mice \cite{pmid22892839} in a population model 
of diet-induced obesity and metabolic disease \cite{pmid29567659}.

The DO mice were derived from eight inbred founder mouse strains, five 
classical lab strains, and three strains more recently derived from 
wild mice \cite{pmid22892839}. They represent three subspecies of mouse 
\textit{Mus musculus domesticus}, \textit{Mus musculus musculus}, and 
\textit{Mus musculus casteneus}, and capture 90\% of the known variation 
in laboratory mice \cite{pmid31133439}. They are maintained with a breeding 
scheme that ensures equal contributions from each founder across the genome
thus rendering almost the whole genome visible to genetic inquiry
\cite{pmid22892839}. We paired clinically relevant metabolic traits
from 500 DO mice [REF], including body weight, plasma levels of insulin
and glucose and plasma lipids, withtranscriptome-wide gene expression 
in four tissues related to metabolic disease: adipose tissue, pancreatic islets, 
liver, and skeletal muscle. Taken together, these data enable a comprehensive 
view into the genetic architecture of metabolic disease.

## Results

### Genetic variation contributed to wide phenotypic variation

Although the environment was consistent across all animals, the 
genetic diversity present in this population resulted in widely 
varying distributions across physiological measurements (Fig.
\ref{fig:trait_overview}). For example, body weights of adult 
individuals varied from less than the average adult B6 body weight 
to several times the body weight of a B6 adult in both sexes 
(Males: `r signif(min(pheno.stats$Body_Weight$Male, na.rm = TRUE), 3)` -
`r signif(max(pheno.stats$Body_Weight$Male, na.rm = TRUE), 3)`g, Females:
`r format(signif(min(pheno.stats$Body_Weight$Female, na.rm = TRUE), 3), nsmall = 1)` -
`r signif(max(pheno.stats$Body_Weight$Female, na.rm = TRUE), 3)`g)
(Fig. \ref{fig:trait_overview}A). Fasting blood glucose (FBG) also
varied considerably (Fig. \ref{fig:trait_overview}B) although 
few of the animals had FBG levels that would indicate pre-diabetes 
(`r as.numeric(pheno.stats$"Prediabetic"["Num_pre-diabetic"])` animals, 
`r as.numeric(pheno.stats$"Prediabetic"["Perc_pre-diabetic"])`\%), 
or diabetes 
(`r as.numeric(pheno.stats$"Diabetic"["Num_diabetic"])` 
animals,
 `r as.numeric(pheno.stats$"Diabetic"["Perc_diabetic"])`\%) according 
 to previously developed cutoffs (pre-diabetes: FBG $\geq$ 
 `r pheno.stats$glucose_ranges[1]` mg/dL, diabetes: FBG $\geq$
 `r pheno.stats$glucose_ranges[2]`, mg/dL) 
\cite{pmid17018838}. Males had higher FBG than females on average 
(Fig. \ref{fig:trait_overview}C) as has been observed before 
suggesting either that males were 
more susceptible to metabolic disease on the high-fat diet, or that 
males and females may require different thresholds for pre-diabetes 
and diabetes. 

Body weight was strongly positively correlated with food consumption 
(Fig. \ref{fig:trait_overview}D $R^2 =$ `r pheno.stats$food_weight["r"]`,
$p<$ `r threshold_p(signif(pheno.stats$food_weight["p"], 2))`) 
and fasting blood glucose (FBG) (Fig. \ref{fig:trait_overview}E, 
$R^2=$ `r pheno.stats$weight_glu["r"]`, $p <$ 
`r threshold_p(signif(pheno.stats$weight_glu["p"], 2))`) 
suggesting a link between behavioral factors and metabolic disease. 
However, the heritability of this trait and others (Fig.
\ref{fig:trait_overview}F) indicates that 
background genetics contribute substantially to correlates of metabolic 
disease in this population.

\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Fig1_trait_overview.pdf} 
\caption{Clinical overview. \textbf{A.} Distributions of final body weight 
in the diversity outbred mice. Sex is indicated by color. The 
average B6 male and female adult weights at 24 weeks of age 
are indicated by blue and green bars on the x-axis. \textbf{B.} The 
distribution of final fasting glucose across the population split 
by sex. Normal, pre-diabetic, and diabetic fasting glucose levels 
for mice are shown by colored bars along the x-axis. \textbf{C.} Males had 
higher fasting blood glucose on average than females. \textbf{D.} The 
relationship between food consumption and body weight for both 
sexes. \textbf{E.} Relationship between body weight and fasting glucose 
for both sexes. \textbf{F.} Heritability estimates for each physiological 
trait. Bars show standard error of the estimate. \textbf{G.} Correlation 
structure between pairs of physiological traits.}
\label{fig:trait_overview}
\end{figure}

The trait correlations (Fig. \ref{fig:trait_overview}G) 
shows that most of the metabolic trait pairs were weakly 
correlated indicating complex relationships among the measured traits. 
This low level of redundancy suggests a broad sampling of multiple 
heritable aspects of metabolic disease including overall body weight, 
glucose homeostasis, pancreatic composition and liver function.

### Distal Heritability Correlated with Phenotype Relevance

```{r, echo = FALSE}
#in the paragraph before a supplemental figure is referenced,
#find its number. The figure counter can be observed at the 
#end to verify the order that the figures should be put in,
#and to see if any figures have not been referenced in the 
#test
fig.label <- "fig:eQTL"
fig.counter <- ref_fig(fig.counter, fig.label)
fig.eqtl.label <- fig.counter[fig.label]
```

We performed eQTL analysis using R/qtl2 \cite{pmid30591514} 
(Methods) and identified both local and distal eQTLs for 
transcripts in each of the four tissues (Supp. Fig `r fig.eqtl.label`). 
Significant local eQTLs far outnumbered distal eQTLs 
(Supp. Fig. `r fig.eqtl.label`F) and tended to be shared across 
tissues (Supp. Fig. `r fig.eqtl.label`G) whereas the few 
significant distal eQTLs we identified tended to be 
tissue-specific (Supp. Fig. `r fig.eqtl.label`H)

We calculated the heritability of each transcript in terms of 
local and distal genetic factors (Methods). Overall, local and 
distal genetic factors contributed approximately equally to 
transcript abundance. In all tissues, both local and distal 
factors explained between `r min.herit` and `r max.herit`\% 
of the variance in the median transcript (Fig \ref{fig:motivation}A). 

\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Fig2_motivation.pdf} 
\caption{Transcript heritability and trait relevance. 
\textbf{A.} Distributions of distal and local heritability of 
transcripts across the four tissues. Overall local and distal 
factors contribute equally to transcript heritability. The 
relationship between (\textbf{B.}) local and (\textbf{C.}) 
distal heritability and trait relevance across all four tissues. 
Here trait relevance is defined as the maximum correlation between 
the transcript and all traits. Local heritability was negatively 
correlated with trait relevance, and distal heritability is 
positively correlated with trait relevance. Pearson ($r$) and $p$ 
values for each correlation are shown in the upper-right of each panel.}
\label{fig:motivation}
\end{figure}

The local heritability of transcripts was negatively correlated with
their trait relevance, defined as the maximum correlation of a 
transcript across all traits (Fig. \ref{fig:motivation}B). This 
suggests that the more local genotype influenced transcript 
abundance, the less effect this variation had on the measured 
traits. Conversely, the distal heritability of transcripts 
was positively correlated with trait relevance (Fig. 
\ref{fig:motivation}C). That is, transcripts that were more 
highly correlated with the measured traits tended to be distally, 
rather than locally, heritable. Importantly, this pattern was 
consistent across all tissues, strongly suggesting that this is a 
generic finding. This finding is consistent with previous observations 
that low-heritability transcripts explain more expression-mediated 
disease heritability than high-heritability transcripts \cite{pmid32424349}. 
However, the positive relationship between trait correlation and distal 
heritability demonstrated further that there are diffuse genetic 
effects throughout the genome converging on trait-related 
transcripts.

### High-Dimensional Mediation identified a high-heritability composite trait that was perfectly mediated by a composite transcript

The above univariate analyses establish the importance of distal 
heritability for trait-relevant transcripts. However, the number 
of transcripts dramatically exceeds the number of degrees of freedom 
of the phenome. Thus, we expect the heritable, trait-relevant transcripts 
to be highly correlated and organized according to coherent, emergent 
biological processes representing the mediating endophenotypes driving 
clinical trait variation. To identify these endophenotypes in a 
theoretically principled way, we developed a novel dimension-reduction 
technique, HDMA, that uses the theory of causal graphical models to 
identify a transcriptomic signature that is simultaneously 1) highly 
heritable, 2) strongly correlated to the measured phenotypes, and 3) 
conforms to the causal mediation hypothesis (Fig. \ref{fig:workflow}).
HDMA projects the high-dimensional scores--a composite genome score 
($G_C$), a composite transcriptome score ($T_C$), and a composite phenome
score ($P_C$)--and uses the univariate theory of mediation to constrain 
these projections to satisfy the hypotheses of perfect mediation. Specifically, 
perfect mediation implies that upon controlling for the transcriptomic score,
the genome score is uncorrelated to the phenome score, which can also be viewed 
as a constraint on the correlation coefficients

\begin{equation*}
Corr(G_C,P_C) = Corr(G_C,T_C)Corr(T_C,P_C),
\end{equation*}

which corresponds to the path coefficient in the mediation model [REF].
Operationally, HDMA is closely related to generalized canonical correlation 
analysis, for which provably convergent algorithms have recently been 
developed \cite{rgcca}. Implementation details for HDMA are available in 
\textbf{Supp. Methods XXX}.

\begin{figure}[ht!]
\includegraphics[width=5in]{Figures/Fig3_workflow.pdf} 
\caption{High-dimensional mediation. \textbf{A.} Workflow indicating 
major steps of high-dimensional mediation. The genotype, transcriptome, 
and phenotype matrices were kernelized to yield single matrices representing 
the relationships between all individuals for each data modality ($G_K$ = 
genome kernel, $T_K$ = transcriptome kernel; $P_K$ = phenome kernel). High-dimensional 
mediation was applied to these matrices to maximize the direct path 
$G \rightarrow T \rightarrow P$, the mediating pathway (arrows), while 
simultaneously minimizing the direct $G \rightarrow P$ pathway (dotted line). 
The composite vectors that resulted from high-dimensional mediation were 
$G_c$, $T_C$, and $P_C$. The partial correlations $\rho$ between these vectors 
indicated perfect mediation. Transcript and trait loadings were calculated 
as described in the methods. \textbf{B.} The null distribution of the path 
coefficient derived from 10,000 permutations compared to the observed path 
coefficient (red line). \textbf{C.} The null distribution of the $G_C$-$T_C$ 
correlation vs. the $T_C$-$P_C$ correlation compared with the observed value 
(red dot).
}
\label{fig:workflow}
\end{figure}

We used high-dimensioal mediation to identify the major axis of variation in 
the transcriptome that mediated the effects of the genome on metabolic traits
(Fig \ref{fig:workflow}). Fig. \ref{fig:workflow}A shows the partial correlations 
($\rho$) between the pairs of these composite vectors. The partial correlation 
between $G_C$ and $T_C$ was `r signif(var.pcor["Causal", "Mediator"], 2)`, 
and the partial correlation between $T_C$ and $P_C$ 
was `r signif(var.pcor["Mediator", "Outcome"], 2)`. However, when the 
transcriptome was taken into account, the partial correlation between 
$G_C$ and $P_C$ was effectively zero (`r signif(var.pcor["Causal", "Outcome"], 2)`).
$P_C$ captured `r signif(comp.trait.exp[[1]], 2)`\% of the overall trait variance,
and its estimated heritability was `r signif(comp.trait.herit[1], 2)` $\pm$ 
`r signif(comp.trait.herit[2], 2)`, which was higher than any of the measured 
traits (Fig. \ref{fig:trait_overview}F). Thus, HDMA identified a maximally 
heritable metabolic composite trait that was perfectly mediated by a highly 
heritable component of the transcriptome.

Standard CCA is prone to over-fitting because in any two large matrices
it can be trivial to identify highly correlated composite vectors [REF]. To
assess whether our implementation of HDMA was similarly prone to over-fitting 
in a high-dimensional space, we performed permutation testing. We permuted 
the individual labels on the transcriptome matrix 1000 times and recalculated 
the path coefficient, which is the partial correlation of $G_C$ and $T_C$ 
multiplied by the partial correlation of $T_C$ and $P_C$. This represents 
the path from $G_C$ to $P_C$ that is mediated through $T_C$. The null 
distribution of the path coefficient is shown in Fig. \ref{fig:workflow}B, 
and the observed path coefficient from the original data is indicated by 
the red line. The observed path coefficient was well outside the null
distribution generated by permutations ($p < 10^{-16}$). Fig. \ref{fig:workflow}C 
illustrates this observation in more detail. Although we identified high 
correlations between $G_C$ and $T_C$, and modest correlations between 
$T_C$ and $P_C$ in the null data (Fig \ref{fig:workflow}C), these two 
values could not be maximized simultaneously in the null data. In contrast,
the red dot shows that in the real data both the $G_C$-$T_C$ correlation 
and the $T_C$-$P_C$ correlation could be maximized simultaneously suggesting 
that the path from genotype to phenotype through transcriptome is highly 
non-trivial and identifiable in this case. These results suggest that these 
composite vectors represent genetically determined variation in phenotype that 
is mediated through genetically determined variation in transcription. 

### Body weight and insulin resistance were highly represented in the expression-mediated composite trait

Each composite score is simply a weighted combination of the measured 
variables and the magnitude and sign of the weights, called loadings, 
correspond the relative importance and directionality of each variable 
in the composite score. The loadings of each measured trait onto $P_C$ 
indicate how much each contributed to the composite phenotype. Final 
body weight contributed the most (Fig. \ref{fig:interpretation}), 
followed by homeostatic insulin resistance (HOMA_IR) and fasting 
plasma insulin levels (Insulin_Fasting). We can thus interpret $P_C$ 
as an index of metabolic disease (Fig. \ref{fig:interpretation}B). 
Individuals with high values of $P_C$ have a higher metabolic index 
and greater metabolic disease, including higher body weight and higher 
insulin resistance. We refer to $P_C$ as the metabolic index going 
forward. Traits contributing the least to the metabolic index were 
measures of cholesterol and pancreas composition. Thus, when we 
interpret the transcriptomic signature identified by HDMA, we are 
explaining primarily the transcriptional mediation of body weight 
and insulin resistance, as opposed to cholesterol measurements.

\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Fig4_interpretation.pdf} 
\caption{Interpretation of loadings. \textbf{A.} Loadings 
across traits. Body weight and insulin resistance contributed 
the most to the composite trait. \textbf{B.} Phenotype scores 
across individuals. Individuals with large positive phenotype 
scores had higher body weight and insulin resistance than average. 
Individuals with large negative phenotype scores had lower body 
weight and insulin resistance than average. \textbf{C.} 
Distribution of transcript loadings in adipose tissue. For 
transcripts with large positive loadings, higher expression was 
associated with higher phenotype scores. For transcripts with 
large negative loadings, higher expression was associated with 
lower phenotype scores. \textbf{D.} Distribution of absolute 
value of transcript loadings across tissues. Transcripts in 
adipose tissue had the largest loadings indicating that 
transcripts in adipose tissue were the best mediators of the 
genetic effects on body weight and insulin resistance.
}
\label{fig:interpretation}
\end{figure}


### High-loading transcripts have low local heritability, high distal heritability, and were linked mechanistically to obesity

We interpreted large loadings onto transcripts as indicating
strong mediation of the effect of genetics on metabolic 
index. Large positive loadings indicate that higher 
expression was associated with a higher metabolic
index (i.e. higher risk of obesity and metabolic disease 
on the high-fat diet) (Fig. \ref{fig:interpretation}C). 
Conversely, large negative loadings indicate that high
expression of these transcripts was associated with a 
lower metabolic index (i.e. lower risk of obesity and 
metabolic disease on the high-fat diet) (Fig. \ref{fig:interpretation}C). 
We used gene set enrichment analysis (GSEA) \cite{fgsea, 
pmid16199517} to look for biological processes and pathways 
that were enriched at the top and bottom of this list (Methods). 

```{r, echo = FALSE}
#in the paragraph before a supplemental figure is referenced,
#find its number. The figure counter can be observed at the 
#end to verify the order that the figures should be put in,
#and to see if any figures have not been referenced in the 
#test
fig.label <- "fig:top_enrich_kegg"
fig.counter <- ref_fig(fig.counter, fig.label)
kegg.enrich.label <- fig.counter[fig.label]
```

In adipose tissue, both GO processes and KEGG pathway 
enrichments pointed to an axis of inflammation and metabolism
(Supp. Fig. `r kegg.enrich.label` and Fig. \ref{fig:top_enrich_go}).
GO terms and KEGG pathways associated with inflammation, particularly 
macrophage infiltration, were positively associated with metabolic
index, indicating that increased expression in inflammatory
pathways was associated with a higher metabolic index. It is well 
established that adipose tissue in obese individuals is 
inflamed [cite] and infiltrated by macrophages [cite], and the 
results here suggest that this may be a heritable component of 
metabolic disease. 

The strongest negative enrichments in adipose tissue were related 
to mitochondial activity in general, and thermogenesis in particular
(Supp. Fig. `r kegg.enrich.label` and Fig. \ref{fig:top_enrich_go}).
It has been shown mouse strains with greater thermogenic potential
are also less susceptible to obesity on a high-fat diet [cite].


```{r, echo = FALSE}
#in the paragraph before a supplemental figure is referenced,
#find its number. The figure counter can be observed at the 
#end to verify the order that the figures should be put in,
#and to see if any figures have not been referenced in the 
#test
fig.label <- "fig:kegg_highlights"
fig.counter <- ref_fig(fig.counter, fig.label)
kegg.highlight.label <- fig.counter[fig.label]
```


Transcripts associated with the citric acid (TCA) cycle as well
as the catabolism of the branched-chain amino acids (BCAA) (valine, 
leuceine, and isoleucine) were strongly enriched with negative 
loadings in adipose tissue (Supp. Fig. `r kegg.highlight.label`). 
Expression of genes in both pathways (for which there is some overlap) 
has been previously associated with insulin sensitivity \cite{pmid29567659, 
pmid22560213, pmid19841271}, suggesting that heritable variation
in regulation of these pathways may influence risk of insulin 
resistance.

Looking a the 10 most positively and negatively loaded 
transcripts from each tissue, it is apparent that transcripts 
in the adipose tissue had the largest loadings, both positive 
and negative, of all tissues (Fig. \ref{fig:loading_heritability}A
bar plot) This suggests that much of the effect of genetics on 
body weight and insulin reisistance is mediated through gene 
expression in adipose tissue. The strongest loadings in liver and 
pancreas were comparable, and those in skeletal muscle were the 
weakest (Fig. \ref{fig:loading_heritability}A), suggesting that 
less of the genetic effects were mediated through transcription 
in skeletal muscle. Heritability analysis showed that transcripts 
with the largest loadings had higher distal 
heritability than local heritability (Fig. 
\ref{fig:loading_heritability}A heat map and box plot). This pattern
contrasts with transcripts nominated by TWAS (Fig. 
\ref{fig:loading_heritability}B), which tended to have lower loadings,
higher local heritability and lower distal heritability. Transcripts
with the highest local heritability in each tissue  (Fig. 
\ref{fig:loading_heritability}C) had the lowest loadings.

We performed a literature search for the genes in each of these groups
along with the terms "diabetes", "obesity", and the name of the 
expressing tissue to determine whether any of these genes had previous 
associations with metabolic disease in the literature (Methods). Multiple 
genes in each group had been previously associated with obesity and 
diabetes (Fig. \ref{fig:loading_heritability} bolded gene names). Genes 
with high loadings were most highly enriched for previous literature 
support. They were `r lit.enrich["loading.twas.enrich"]` more likely 
than TWAS hits and `r lit.enrich["loading.herit.enrich"]` times more 
likely than genes with high local heritability to be previously 
associated with obesity or diabetes.

\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Fig5_loading_heritability.pdf} 
\caption{Transcripts with high loadings have high distal heritability
and literature support. Each panel has a bar plot showing the loadings 
of transcripts selected by different criteria. Bar color indicates the 
tissue of origin. The heat map shows the local (L - left) and distal 
(D - right) heritability of each transcript. \textbf{A.} Loadings for 
the 10 transcripts with the largest positive loadings and the 10 
transcripts with the largest negative loadings for each tissue. 
\textbf{B.} Loadings of TWAS candidates with the 10 largest positive 
correlations with traits and the largest negative correlations with 
traits across all four tissues. \textbf{C.} The transcripts with the 
largest local heritability (top 20) across all four tissues.
}
\label{fig:loading_heritability}
\end{figure}

### Tissue-specific transriptional programs were associated with metabolic traits

Clustering of transcripts with top loadings in each tissue showed 
tissue-specific functional modules associated with obesity and 
insulin resistance (Fig. \ref{fig:toa}A) (Methods). 
The clustering highlights the importance of immune activation 
particularly in adipose tissue. Except for the "mitosis" cluster,
which had large positive loadings in three of the four tissues,
all clusters were strongly loaded in only one or two tissues. 
For example, the lipid metabolism cluster was loaded most heavily
in liver. The positive loadings suggest that high expression of
these genes particularly in the liver was associated with increased
metabolic disease. This cluster included the gene \textit{Pparg}, 
whose primary role is in the adipose tissue where it is considered 
a master regulator of adipogenesis \cite{pmid17389767}. Agonists of 
\textit{Pparg}, such as thiazolidinediones, are FDA-approved to 
treat type II diabetes, and reduce inflammation and adipose 
hyptertrophy \cite{pmid17389767}. Consistent with this role, the loading for 
\textit{Pparg} in adipose tissue was negative, suggesting 
that higher expression was associated with leaner mice (Fig. \ref{fig:toa}B). 
In contrast, \textit{Pparg} had a large positive loading in liver, 
where it is known to play a role in the development of hepatic steatosis, 
or fatty liver. Mice that lack \textit{Pparg} specifically in the liver, 
are protected from developing steatosis and show reduced expression
of lipogenic genes \cite{pmid12805374, pmid12618528}. Overexpression
of \textit{Pparg} in the livers of mice with a \textit{Ppara} knockout, 
causes upregulation of genes involved in adipogenesis \cite{pmid16357043}.
In the livers of both mice and humans
high \textit{Pparg} expression is associated with hepatocytes that
accumulate large lipid droplets and have gene expression profiles 
similar to adipocytes \cite{pmid15644454, pmid16403437}.

The local and distal heritability of \textit{Pparg} is low in adipose 
tissue suggesting its expression in this tissue is highly constrained 
in the population (Fig. \ref{fig:toa}B). However, the distal heritability 
of \textit{Pparg} in liver is relatively high suggesting it is complexly 
regulated and has sufficient variation in this population to drive 
variation in phenotype. Both local and distal heribatility of 
\textit{Pparg} in the islet are relatively high, but the loading is low,
suggesting that variability of expression in the islet does not
drive variation in metabolic index. These results highlight the importance
of tissue context when investigating the role of heritable transcript
variability in driving phenotype.

Gene lists for all clusters are available in Supplemental File XXX.

\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Fig6_TOA.pdf} 
\caption{Tissue-specific transcriptional programs were associated 
with obesity and insulin resistance. \textbf{A} Heat map showing 
the loadings of all transcripts with loadings greater than 2.5 
standard deviations from the mean in any tissue. The heat map was 
clustered using k medoid clustering. Functional enrichments of each 
cluster are indicated along the left margin. \textbf{B} Loadings for 
\textit{Pparg} in different tissues. \textbf{C} Local and distal of 
\textit{Pparg} expression in different tissues.
}
\label{fig:toa}
\end{figure}


### Gene expression, but not local eQTLs, predicted body weight in an independent population

To test whether the transcript loadings identified in the DO could 
be translated to another population, we tested whether they could 
predict metabolic phenotype in an independent population of
CC-RIX mice, which were F1 mice derived from multiple pairings of 
Collaborative Cross (CC) [cite] strains (Fig. \ref{fig:cc_prediction}) 
(Methods). We tested two questions. First, we asked whether the 
loadings identified in the DO mice were relevant to the relationship 
between the transcriptome and the phenome in the CC-RIX. We predicted
body weight (a surrogate for metabolic index) in each CC-RIX individual 
using measured gene expression in each tissue and the transcript loadings 
identified in the DO (Methods). The predicted body weight and acutal body 
weight were highly correlated in all tissues (Fig. \ref{fig:cc_prediction}B 
left column). The best prediction was achieved for adipose tissue, which 
supports the observation in the DO that adipose expression was the strongest 
mediator of the genetic effect on metabolic index. This result also confirms 
the validity and translatability of the transcript loadings and their relationship 
to metabolic disease.

\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Fig7_CC_Prediction.pdf} 
\caption{Transcription, but not local genotype, predicts 
phenotype in the CC-RIX. \textbf{A.} Workflow showing procedure 
for translating HDMA results to an independent population of mice. 
\textbf{B.} Relationships between the predicted metabolic index 
and measured body weight. The left column shows the predictions 
using measured transcripts. The right column shows the prediction 
using transcript levels imputed from local genotype. Gray boxes 
indicate measured quantities, and blue boxes indicate calculated 
quantities. The dots in each panel represent individual CC-RIX strains. 
The gray lines show the standard deviation on body weight for the strain.
}
\label{fig:cc_prediction}
\end{figure}


```{r, echo = FALSE}
#in the paragraph before a supplemental figure is referenced,
#find its number. The figure counter can be observed at the 
#end to verify the order that the figures should be put in,
#and to see if any figures have not been referenced in the 
#test
fig.label <- "fig:cc_imputation"
fig.counter <- ref_fig(fig.counter, fig.label)
cc.imputation.label <- fig.counter[fig.label]
```


The second question related to the source of the relevant variation 
in gene expression. If local regulation was the predominant factor 
influencing gene expression, we should be able to predict phenotype 
in the CC-RIX using transcripts imputed from local genotype 
(Fig. \ref{fig:cc_prediction}A). The DO and the CC-RIX were derived 
from the same eight founder strains and so carry the same alleles 
throughout the genome. We imputed gene expression in the CC-RIX
using local genotype and were able to estimate variation in gene 
transcription robustly (Supp. Fig. `r cc.imputation.label`).
However, these imputed values failed to predict body weight in 
the CC-RIX when weighted with the loadings from HDMA. (Fig. 
\ref{fig:cc_prediction}B right column). This result suggests 
that local regulation of gene expression is not the primary 
factor driving heritability of complex traits, consistent 
with our findings in the DO population that distal heritability 
was a major driver of trait-relevant variation and that high-loading 
transcripts had comparatively high distal and low local heritability.

### Distally heritable transcriptomic signatures reflected variation in composition of adipose tissue and islets

The interpretation of global genetic influences on gene expression and 
phenotype is potentially more challenging than the interpretation and 
translation of local genetic influences, as genetic effects cannot 
be localized to individual gene variants or transcripts. However, 
there are global patterns across the loadings that can inform 
mechanism. For example, heritable variation in cell type 
composition can be derived from transcript loadings. We observed
above that immune activation in the adipose tissues was an 
important driver of obesity in the DO population. To determine
whether this is reflected as an increase in macrophages in adipose
tissue, we compared loadings of cell-type specific genes in adipose
tissue (Methods). The mean loading of macrophage-specific genes was 
significantly greater than 0 (Fig. \ref{fig:human_translation}A), 
indicating that obese mice were genetically predisposed to have high 
levels of macrophage infiltration in adipose tissue in response to 
the high-fat, high-sugar diet. 

\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Fig8_Human_Translation.pdf} 
\caption{HDMA results translate to humans. \textbf{A.} Distribution of 
loadings for cell-type-specific transcripts in adipose tissue. \textbf{B.} 
Distribution of loadings for cell-type-specific transcripts in pancreatic 
islets (green). \textbf{C.} Null distributions for the mean loading of 
randomly selected transcripts in each cell type compared with the observed 
mean loading of each group of transcripts (red asterisk). \textbf{D.} 
Predictions of metabolic phenotypes in four adipose transcription data 
sets downloaded from GEO. In each study the obese/diabetic patients were 
predicted to have greater metabolic disease than the lean/non-diabetic 
patients based on the HDMA results from DO mice.
}
\label{fig:human_translation}
\end{figure}

We also compared loadings of cell-type specific transcripts in islet 
(Methods). The mean loadings for alpha-cell specific transcripts were 
significantly greater than 0, while the mean loadings for delta- and 
endothelial-cell specific genes were significantly less than 0
(Fig. \ref{fig:human_translation}B). These results suggest either 
that mice with higher metabolic index had inherited a higher proportions 
of alpha cells, and lower proportions of endothelial and delta cells 
in their pancreatic islets, that such compositional changes were induced
by the HFHS diet in a heritable way, or both. In either case, these 
results support the hypothesis that alterations in islet composition 
drive variation in metabolic index.

Notably, the loadings for pancreatic beta cell-type specific loadings 
was not significantly different from zero. This is not necessarily 
reflective of the function of the beta cells in the obese mice, 
but rather suggests that any variation in the number of beta cells 
in these mice was unrelated to obesity and insulin resistance. This
is further consistent with the islet composition traits having small
loadings in the phenome score (Fig. \ref{fig:interpretation}).

### Heritable transcriptomic signatures translated to human disease

Ultimately, the heritable transcriptomic signatures that we 
identified in DO mice will be useful if they inform
pathogenicity and treatment of human disease. To investigate 
the potential for translation of the gene signatures identified
in DO mice, we compared them to transcriptional profiles in 
obese and non-obese human subjects (Methods). We limited our 
analysis to adipose tissue because the adipose tissue signature 
had the strongest relationship to obesity and insulin resistance
in the DO. 

We calculated a predicted obesity score for each individual 
in the human studies based on their adipose tissue gene
expression (Methods) and compared the predicted scores for 
obese and non-obese groups as well as diabetic and non-diabetic
groups. In all cases, the predicted obesity scores were higher
on average for individuals in the obese and diabetic groups
compared with the lean and non-diabetic groups (Fig. 
\ref{fig:human_translation}D). This indicates that the distally 
heritable signature of obesity identified in DO mice is relevant 
to obesity and diabetes in human subjects.

### Targeting gene signatures
Another potential application of the transcript loading landscape 
is in ranking potential drug candidates for the treatment of 
metabolic disease. Although high-loading transcripts may be 
good candidates for understanding specific biology related to 
obesity, the transcriptome overall is highly interconnected and 
redundant, and focusing on individual transcripts for treatment 
may be less effective than using broader transcriptomic signatures 
that capture the emergent biology [cite or remove]. The ConnectivityMap 
(CMAP) database \cite{pmid17008526} developed by the Broad Institute 
allows us to query thousands of compounds that reverse or enhance the 
extreme ends of transcriptomic signatures in multiple different cell 
types. By identifying drugs that reverse pathogenic transcriptomic 
signatures, we can potentially identify compounds that have favorable 
effects on gene expression.

To test this hypothesis, we queried the CMAP database through the 
CLUE online query tool (\url{https://clue.io/query/}, version 
1.1.1.43) (Methods). We identified top anti-correlated hits 
across all cell types. To get more tissue-specific results, we
also looked at top results in cell types that most closely 
resembled our tissues. We looked at results in adipocytes (ASC)
as well as pancreatic tumor cells (YAPC) regardless of $p$ value 
(Supplemental Figure XXX and XXX). 

Looking broadly across cell types, the notable top hits from the 
adipose tissue loadings included mTOR inhibitors and glucocorticoid 
agonists (Supplemental Figure XXX). It is thought that metformin, 
which is commonly used to improve glycemic control, acts, at least 
in part, by inhibiting mTOR signaling \cite{pmid30290005, 
pmid30034573}. However, long-term use of other mTOR inhibitors, 
such as rapamycin, are known to cause insulin resistance and 
$\beta$-cell toxicity \cite{pmid30034573, pmid23881200, pmid21266327}. 
Glucocorticoids are used to reduce inflammation, which was a 
prominent siganture in the adipose tissues, but these drugs also 
promote hyperglycemia and diabetes \cite{pmid24582093, pmid35585199}.
Accute treatment with glucocorticoids has further been shown to 
reduce thermogenesis in rodent adipocytes \cite{pmid30310815, 
pmid11254472, pmid23197361}, but increase thermogenesis in human 
adipocytes \cite{pmid27411014, pmid25385872}. Thus, the pathways 
identified by CMAP across all cell types were highly related to 
the transcript loading profiles, but the relationship was not a 
simple reversal.

The top hit for the adipose composite transcript in CMAP adipocytes 
was a PARP inhibitor (Supplemental Figure XXXB). PARPs play a role 
in lipid metabolism and are involved in the development of obesity 
and diabetes \cite{pmid34450194}. PARP1 inhibition increases 
mitochondrial biogenesis \cite{pmid21459330}. Inihibition of PARP1 
activity can further prevent necrosis in favor of the less 
inflammatory apoptosis \cite{pmid12114611}, thereby potentially 
reducing inflammation in stressed adipocytes. Other notable hits among 
the top 20 were BTK inhibitors, which have been observed to suppress 
inflammation and improve insulin resistance \cite{pmid33648925} as well 
as to reduce insulin antibodies in type I diabetes \cite{pmid28753229}. 
IKK inhibitors have been shown to improve glucose control in type II
diabetes \cite{pmid28683283, pmid15685170}. 

Among the top most significant hits for the transcript loadings from 
pancreatic islets (Fig. XXX), was suppression of T cell receptor 
signaling, which is known to be involved in Type 1 diabetes 
\cite{pmid33603744}, as well as TNFR1, which has been associated with 
mortality in diabetes patients \cite{pmid32281000}. Suppression of 
NOD1/2 signaling was also among the top hits. NOD1 and 2 sense ER 
stress \cite{pmid27007849, pmid28823510}, which is associated with 
$\beta$-cell death in type 1 and type 2 diabetes \cite{pmid24520198}. 
This cell death process is dependent on NOD1/2 signaling 
\cite{pmid27007849}, although the specifics have not yet been worked 
out.

We also looked specifically at hits in pancreatic tumor cells (YAPC) 
regardless of significance level. Hits in this list included widely 
used diabetes drugs, such as sulfonylureas, PPAR receptor agonists, 
and insulin sensitizers. Rosiglitazone is a PPAR-$\gamma$ agonist 
and was one of the most prescribed drugs for type 2 diabetes before 
its use was reduced due to cardiac side-effects \cite{pmid21190462}. 
Sulfonylureas are another commonly prescribed drug class for type 2 
diabetes, but also have notable side effects including hypoglycemia 
and accellerated $\beta$-cell death \cite{pmid16631807}.


## Discussion

Here we used a novel high-dimensional mediation analysis (HDMA)
to investigate the relative contributions of local and distal
gene regulation to heritable trait variation in a genetically 
diverse mouse model of diet-induced obesity and metabolic 
disease. We identified tissue-specific composite transcripts 
mediating the effect of genetic background on metabolic traits.
Transcripts contributing most strongly to these composite
transcripts were distally heritable. These composite transcripts, 
but not local eQTL, were able to predict obesity in an independent
mouse population with divergent allelic structure. Moreover,
the composite transcript from adipose tissue predicted obesity 
and diabetes status in human cohorts with measured adipose gene 
expression. Taken together, these results support the hypothesis 
that gene expression mediating the effect of genetic background 
on phenotype are primarily distally regulated, and that the gene 
regulatory networks influencing metabolic disease are conserved 
across mice and humans.

It has frequently been assumed that gene regulation in 
\textit{cis} is the primary driver of genetically associated 
trait variation, but attempts to use local gene regulation 
to explain phenotypic variation have had limited success 
\cite{pmid32912663, pmid36515579}. In recent years, evidence
has mounted that distal gene regulation may be an important
mediator of trait heritability \cite{pmid32424349, 
pmid37857933, pmid31051098}. It has been observed
that transcripts with high local heritability explain less
expression-mediated disease heritability than transcripts
with low local heritability \cite{pmid32424349}. Consistent with
this observation, genes located near GWAS hits tend to be 
complexly regulated \cite{pmid37857933}. They also tend to be
enriched with functional annotations, in contrast to genes with
simple local regulation, which tend to be depleted of functional 
annotations suggesting they are less likely to be directly involved 
in disease traits \cite{pmid37857933}. These ideas are consistent 
with principles of robustness in complex systems \cite{pmid29782925, 
pmid12082173, pmid27304973}. If a transcript were both important to 
a trait and subject to strong local regulation, a population would 
be susceptible to extremes in phenotype that might frequently cross 
the threshold to disease. Indeed, strong disruption of highly 
trait-relevant genes is the cause of Mendelian disease. 

Our findings were consistent with previous observations. The 
composite transcripts we identified were highly heritable and 
explained a high proportion of disease risk. Transcript 
loadings (the degree to which they contributed to the composite
transcript) were negatively correlated with local heritability 
and positively correlated with distal heritability. These 
strongly loaded transcripts were also enriched for functional 
annotations associated with metabolic disease. The composite 
transcripts were moreover able to predict obesity in an independent 
cohort of mice whereas models using local eQTL only could not. 
Together these observations suggest that distal gene regulation 
was the dominant mode through which gene expression mediated the 
effect of genetic background on complex traits. 

Identification of this distally heritable signature depended
on the high-dimensional approach we used. Because HDMA uses a 
kinship matrix rather than genotypes at individual loci, it 
allows for arbitrarily complex gene regulation, as well as 
the interconnectedness and redundancy of the transcriptome. 
This feature also means that HDMA assumes that traits are 
highly polygenic. It also assumes that genetic effects are 
distributed across the genome. In contrast, one-dimensional,
univariate approaches assume a large, localized genetic effect. 
Thus, the HDMA approach is consistent with the omnigenic model 
of complex traits which posits that complex traits are massively 
polygenic and that their heritability is spread out across the 
genome \cite{pmid28622505}. In the omnigenic model, genes are 
classified either as "core genes," which directly impinge on the 
trait, or "peripheral genes," which are not directly trait-related, 
but influence core genes through the complex gene regulatory network. 
HDMA explicitly models a central proposal of the omnigenic model which 
posits that once the expression of the core genes (i.e. trait-mediating 
genes) is accounted for, there should be no residual correlation between 
the genome and the phenome. Here, when the composite transcript
was taken into account there was no residual correlation between
the composite genome and composite phenome (Fig. \ref{fig:workflow}A). 

The composite transcript is essentially a weighted vector with larger 
weights (loadings) indicating higher "core-ness" of a transcript. There 
was no clear demarcation between the core and peripheral genes in loading
magnitude, but we do not necessarily expect a clear separation given the 
complexity of gene regulation and the genotype-phenotype map 
\cite{pmid29906445}. Still, the transcripts with the largest loadings
had high distal heritability, low local heritability, and were enriched 
for biological processes related to metabolic traits, as we would predict
for core genes.


```{r, echo = FALSE}
#in the paragraph before a supplemental figure is referenced,
#find its number. The figure counter can be observed at the 
#end to verify the order that the figures should be put in,
#and to see if any figures have not been referenced in the 
#test
fig.label <- "fig:Nucb2_eqtl"
fig.counter <- ref_fig(fig.counter, fig.label)
nucb2.eqtl.label <- fig.counter[fig.label]
```

An extension of the omnigenic model proposed that most heritability
of complex traits is driven by weak distal eQTLs \cite{pmid31051098}.
This is consistent with what we observed here. The transcripts with 
the largest loadings were strongly distally regulated and only weakly 
locally regulated, suggesting that distal gene regulation plays a 
primary role in driving heritable trait variation. We saw further 
that the patterns of distal heritability were complex spread across 
the genome. Even for transcripts whose expression was strongly 
regulated by distal factors, these factors were multiple and 
spread across the genome. For example, \textit{Nucb2}, was a 
strongly mediating transcript in islet and was also strongly 
distally regulated (66\% distal heritability) (Fig. 
\ref{fig:loading_heritability}). This gene is expressed in pancreatic 
$\beta$ cells and is involved in insulin and glucagon release 
\cite{pmid22108805, pmid23537085, pmid24993278}. Although its 
transcription was highly heritable in islets, that regulation was 
distributed across the genome, with no clear distal eQTL 
(Supp. Fig. `r nucb2.eqtl.label`). Thus, although distal regulation 
of some genes may be strong, this regulation is likely to be highly 
complex and not easily localized. 

The composite transcripts identified by HDMA are richly 
interpretable in both tissue- and gene-specific manners. 
The transcripts with the strongest loadings were enriched 
in biological functions previously known to be involved 
in the pathogenesis of metabolic disease, such as inflammation 
in adipose tissue. That these processes were identified in 
this analysis suggests that they have a heritable component, 
and that some individuals are genetically susceptible to greater
adipose inflammation on a high-fat, high-sugar diet.

Individual transcripts also demonstrated biologically 
interpretable, tissue-specific patterns. We highlighted
\textit{Pparg}, which is known to be protective in adipose 
tissue \cite{pmid17389767} where it was negatively loaded, 
and harmful in the liver \cite{pmid12805374, pmid12618528, 
pmid16357043, pmid15644454, pmid16403437}, where it was 
positively loaded. Such granular patterns may be useful
in generating hypotheses for further testing, and 
prioritizing genes as therapeutic targets. The 
tissue-specific nature of the loadings also may provide
clues to tissue-specific effects, or side effects, 
of targeting particular genes system-wide, since 
antagonists of \textit{Pparg} may reduce fatty 
liver disease, but exacerbate adipose tissue inflammation.

We showed further that these composite transcripts can 
be used as weighted vectors in multiple types of analysis, 
such as drug prioritization using gene set enrichment 
analysis (GSEA) and the CMAP database. In particular,
the CMAP analysis identified drugs which have geen 
demonstrated to reverse insulin resistance and other 
aspects of metabolic disease. This finding supports the 
causal role of these gene signatures in pathogenesis of
metabolic disease and thus their utility in prioritizing 
drugs and gene targets as therapeutics.

Another useful appllication of the composite transcripts is
to pair them with cell-type specific genes to generate hypotheses 
about cell composition in individual tissues. Combining the 
multi-tissue, transcriptome-wide weighted vectors with public 
databases and data sets thus provides a path for generating a 
wide range of testable hypotheses. Moreover, each data set 
presented here was derived from human tissues or cell lines,
thus demonstrating the translatability of these results. That 
the mouse-derived adipose composite transcript was able to 
classify human adipose tissue in terms of obesity and diabetes 
status further supports the direct translatablility of these 
findings, utility of HDMA, and the continued importance of 
mouse models of human disease in which it is possible to 
obtain complete transriptomes in mutliple tissues across
large numbers of individuals. 
 
In conclusion, we have shown that both tissue specificity and 
distal gene regulation are critically important to understanding 
the genetic architecture of complex traits. We identified important 
genes and gene signatures that were heritable, causal of disease, 
and translatable to other mouse populations and to humans. Finally, 
we have shown that by directly acknowledging the complexity of both 
gene regulation and the genotype-to-phenotype map, we can gain a new 
perspective on disease pathogenesis and develop actionable hypotheses 
about pathogenic mechanisms and potential treatments. 

## Data Availability
Here we tell people where to find the data

## Acknowledgements
Here we thank people

\pagebreak

## Supplemental Figures

\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Supp_Fig1_eQTL.pdf} 
\caption{Overview of eQTL analysis in DO mice. \textbf{A.} RNA seq 
samples from the four different tissues clustered by tissue. 
\textbf{B.-E.} eQTL maps are shown for each tissue. The $x$-axis 
shows the position of the mapped eQTL, and the $y$-axis shows the 
physical position of the gene encoding each mapped transcript. 
Each dot represents an eQTL with a minimum LOD score of 8. The dots 
on the diagonal are locally regulated eQTL for which the mapped eQTL 
is at the within 4Mb of the encoding gene. Dots off the diagonal are 
distally regulated eQTL for which the mapped eQTL is distant from the 
gene encoding the transcript. \textbf{F.} Comparison of the total number 
of local and distal eQTL with a minimum LOD score of 8 in each tissue. 
All tissues have comparable numbers of eQTL. Local eQTL are much more 
numerous than distal eQTL. \textbf{G.} Counts of transcripts with local 
eQTL shared across multiple tissues. The majority of local eQTL were 
shared across all four tissues. \textbf{H.} Counts of transcripts with 
distal eQTL shared across multiple tissues. The majority of distal eQTL 
were tissue-specific and not shared across multiple tissues. For both G 
and H, eQTL for a given transcript were considered shared in two tissues 
if they were within 4Mb of each other. Colored bars indicate the counts 
for individual tissues for easy of visualization.
}
\label{fig:eQTL}
\end{figure}


\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Supp_Fig_enrichments_KEGG.pdf} 
\caption{Bar plots showing normalized enrichment scores (NES) for KEGG 
pathways as determined by fast gene score enrichment analysis (fgsea). 
Only the top 10 positive and top 10 negative scores are shown. Colors 
indicate tissue. The name beside each bar shows the name of each enriched 
KEGG pathway.
}
\label{fig:top_enrich_kegg}
\end{figure}

\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Supp_Fig_enrichments_GO.pdf} 
\caption{Bar plots showing normalized enrichment scores (NES) for GO 
terms as determined by fast gene score enrichment analysis (fgsea). 
Only the top 10 positive and top 10 negative scores are shown. Colors 
indicate tissue. The name beside each bar shows the name of each enriched 
GO term. The letters in parentheses indicate whether the term is from the 
biological process ontology (BP), the molecular function ontology (MF), 
or the cellular compartment ontology (CC).
}
\label{fig:top_enrich_go}
\end{figure}


\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Supp_Fig_CC-RIX_Imputation.pdf} 
\caption{Validation of transcript imputation in the CC-RIX. \textbf{A.} 
Distributions of correlations between imputed and measured transcripts 
in the CC-RIX. The mean of each distribution is shown by the red line. 
All distributions were skewed toward positive correlations and had
 positive means near a Pearson correlation (r) of 0.5. \textbf{B.} 
 The relationship between the correlation between measured and 
 imputed expression in the CC-RIX (x-axis) and eQTL LOD score. As 
 expected, imputations are more accurate for transcripts with strong 
 local eQTL. \textbf{C.} Variance explained by local genotype in the 
 DO and CC-RIX. 
}
\label{fig:cc_imputation}
\end{figure}


\begin{figure}[ht!]
\includegraphics[width=\textwidth]{Figures/Supplemental_FigX_Nucb2_eQTL.pdf} 
\caption{Regulation of \textit{Nucb2} expression in islet. \textit{Nucb2} 
is encoded on mouse chromosome 7 at 116.5 Mb (red line). In islets the 
heritability of \textit{Nucb2} expression levels is 69\% heritable. This 
LOD score trace shows that there is no local eQTL at that position, nor 
any strong distal eQTL anywhere else in the genome. 
}
\label{fig:Nucb2_eqtl}
\end{figure}

\medskip

\bibliographystyle{unsrt}
\bibliography{islet.bib}
